---
title: "Analysis of data"
author: "Emilia König et al."
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

## Libraries ------------

library(vegan)
library(tidyverse)
library(janitor)
library(readxl)
library(glue)
library(viridis)
library(lme4)
library(nlme)
library(here)

theme_set(theme_bw(base_size = 16))
```

# Data overview

```{r, echo=FALSE, message=FALSE, warning=FALSE}
## Data loading --------

here::i_am(path = "R/AnalysisSummary_15-09-2022.Rmd")

data_path <- here("data/")
data_files <- list.files(path = data_path)


## read data to data.frames
biom_nonfil_df <- read_csv(file = here(data_path, data_files[1]))
metadata_df <- readxl::read_xlsx(path = here(data_path, data_files[2]), sheet = 1)
secuencias_tree_df <- read_csv(file = here(data_path, data_files[3]))
tax_table_nonfil_df <- read_csv(file = here(data_path, data_files[4]))

# add a name to un-named first columns
names(biom_nonfil_df)[1] <- names(secuencias_tree_df)[1] <- names(tax_table_nonfil_df)[1] <- "code"

```

The data has `r nrow(biom_nonfil_df)` rows and `r ncol(biom_nonfil_df)` columns.


# Alpha diversity

## Shannon index

Shannon index values were calculated with the `vegan` package function `diversity`.

```{r, echo=FALSE}
##########
### Calculating the Shannon index ---------------
##########

# Calculate Shannon index from pig data
biom_shannon <- 
  biom_nonfil_df %>% 
  # removing "non-pig" data
  select(-code, -`CN_MCC21-111_S106`) %>% 
  # Calculate the Shannon index for columns (i.e. data for all pigs)
  diversity(MARGIN = 2, index = "shannon")

## add data to a data.frame
shannon_df <- tibble("sample_id" = names(biom_shannon), "shannon" = biom_shannon) %>% 
  ## add information of manipulator pigs etc
  left_join(select(metadata_df, SAMPLE_ID, MANIPULATOR, `PAIR NO.`, SEX, SIZE) %>% 
              # rename columns to match existing names
              rename(sample_id = SAMPLE_ID, manipulator = MANIPULATOR, pair = `PAIR NO.`, sex = SEX, size = SIZE),
            # combine by sample id
            by = "sample_id") %>% 
  # Change manipulator to factor and change level names
  mutate(status = if_else(manipulator == 1, true = "Manipulator", false = "Control") %>% factor(levels = c("Manipulator", "Control")),
         sex = if_else(sex == 1, true = "female", false = "barrow") %>% factor(levels = c("female", "barrow")))


######
##  Create visualization -------------------
######

shannon_df %>% 
  ggplot(aes(x = status, y = shannon)) +
  geom_boxplot(fill = "grey", outlier.size = 0, alpha = 0.5) +
  geom_jitter(aes(color = sex), width = 0.1, size = 2) +
  scale_y_continuous(limits = c(4,6), breaks = c(4,4.5,5,5.5,6)) +
  labs(x = "", y = "Shannon index", color = "Sex") +
  theme(legend.position = "top") +
  scale_color_brewer(palette = "Set1")

```



```{r, echo=FALSE}
glmfit <- glm(shannon ~ status*sex, data = shannon_df)
summary(glmfit)
plot(glmfit)
```

A general linear model was used for evaluating the effects of `status` and `sex` on the calculated Shannon index values. The `status` was not a statistically significant term in the model with a $p$-value of `r broom::tidy(glmfit) %>% filter(term == "statusControl") %>% pull(p.value) %>% round(digits = 2)`.

```{r}
wilcox.test(shannon ~ status, data = shannon_df, conf.int = TRUE) 
```

The nonparametric Mann-Whitney test is quite close to being statistically significant at $\alpha$-value `0.05`.

### Log-transformed Shannon

```{r, echo=FALSE}
shannon_df <- shannon_df %>% 
  mutate(ln_shannon = log(shannon))

glmfit_log <- glm(ln_shannon ~ status*sex, data = shannon_df)
summary(glmfit_log)
plot(glmfit_log)
```


## Chao1 Index

Richness was evaluated by calculating the Chao1 index using the `estimateR` function from the `vegan` package.

```{r, echo=FALSE}
# Calculate richness from pig data
biom_richness <- 
  biom_nonfil_df %>% 
  # removing "non-pig" data
  select(-code, -`CN_MCC21-111_S106`) %>% 
  # Calculate the Shannon index for columns (i.e. data for all pigs)
  apply(MARGIN = 2, function(x){sum(x>0)})

## Calculate Chao1
biom_chao1 <- 
  biom_nonfil_df %>% 
  # removing "non-pig" data
  select(-code, -`CN_MCC21-111_S106`) %>% 
  # Calculate the Shannon index for columns (i.e. data for all pigs)
  apply(MARGIN = 2, function(x){estimateR(x)}) %>% 
  .["S.chao1",]


shannon_df <- shannon_df %>% 
  left_join(y = tibble("richness" = biom_richness, "chao1" = biom_chao1, sample_id = names(biom_richness)), by = "sample_id")


shannon_df %>% 
  ggplot(aes(x = status, y = chao1)) +
  geom_boxplot(fill = "grey", outlier.size = 0, alpha = 0.5) +
  geom_jitter(aes(color = sex), width = 0.1, size = 2) +
  #scale_y_continuous(limits = c(4,6), breaks = c(4,4.5,5,5.5,6)) +
  labs(x = "", y = "Chao1 index", color = "Sex") +
  theme(legend.position = "top") +
  scale_color_brewer(palette = "Set1")
```




```{r, echo=FALSE}

glmfit2 <- glm(chao1 ~ status*sex, data = shannon_df)
summary(glmfit2)
plot(glmfit2)
```

A general linear model was used for evaluating the effects of `status` and `sex` on the calculated Chao1 index values. The `status` was not a statistically significant term in the model with a $p$-value of `r broom::tidy(glmfit2) %>% filter(term == "statusControl") %>% pull(p.value) %>% round(digits = 2)`.

```{r}
wilcox.test(chao1 ~ status, data = shannon_df, conf.int = TRUE) 
```

The nonparametric Mann-Whitney test is not statistically significant.

# Beta-diversity

```{r, echo=FALSE}
#########
## Principal Coordinate analysis ------------
#########

# calculating Bray-Curtis distances
biom_dist <- 
  biom_nonfil_df %>% 
  # removing "non-pig" data
  select(-code, -`CN_MCC21-111_S106`) %>% 
  # transpose the data
  t() %>% 
  vegdist()

# first two axes in PCoA
PCoA_res <- cmdscale(biom_dist, eig = TRUE, add = TRUE)
positions <- PCoA_res$points


colnames(positions) <- c("PCoA1", "PCoA2")
positions <- positions %>% 
  as_tibble(rownames = "sample_id") %>% 
  left_join(y = metadata_df %>% 
              select(SAMPLE_ID, MANIPULATOR, `PAIR NO.`) %>% 
              rename(sample_id = SAMPLE_ID, manipulator = MANIPULATOR, pair = `PAIR NO.`),
            by = "sample_id") %>% 
  mutate(manipulator = if_else(manipulator == 1, true = "Manipulator", false = "Control") %>% as_factor())

#calculate percent explained
percents <- 100*PCoA_res$eig / sum(PCoA_res$eig)
percents <- round(percents, digits = 1)

lab_percent <- c(glue("PCoA 1({percents[1]}%)"), glue("PCoA 2({percents[2]}%)"))

## Visualization with connecting lines ---

controls <- positions %>% 
  filter(manipulator == "Control")

manipulators <- positions %>% 
  filter(manipulator == "Manipulator")


controls <- controls %>% 
  # add manipulator pairs and their PCoA coordinates as end points for connecting lines
  left_join(select(manipulators, -sample_id, -manipulator) %>% rename(xend = PCoA1, yend = PCoA2), by = "pair")

manipulators <- manipulators %>% 
  # add manipulator pairs and their PCoA coordinates as end points for connecting lines
  left_join(select(controls, -sample_id, -manipulator, -xend, -yend) %>% rename(xend = PCoA1, yend = PCoA2), by = "pair")

positions2 <- rbind(controls, manipulators)

positions2 %>% 
  #let's switch factor levels to get same coloring
  mutate(manipulator = factor(manipulator, levels = c("Manipulator", "Control"))) %>% 
  ggplot(aes(x = PCoA1, y = PCoA2, color = manipulator)) +
  geom_point(size = 2, alpha = 0.8) +
  labs(x = lab_percent[1], y = lab_percent[2], color = "") +
  theme(legend.position = "top") +
  geom_text(aes(x = PCoA1+0.02, label = pair), color = "black", alpha = 0.8) +
  geom_segment(aes(xend = xend, yend = yend), color = "grey", alpha = 0.6) +
  scale_colour_brewer(palette = "Set1")


positions2 %>% 
  #let's switch factor levels to get same coloring
  mutate(manipulator = factor(manipulator, levels = c("Manipulator", "Control"))) %>% 
  ggplot(aes(x = PCoA1, y = PCoA2, color = manipulator)) +
  geom_point(size = 2, alpha = 0.8) +
  labs(x = lab_percent[1], y = lab_percent[2], color = "") +
  theme(legend.position = "top") +
  geom_segment(aes(xend = xend, yend = yend), color = "grey", alpha = 0.6) +
  scale_colour_brewer(palette = "Set1")


```


## PERMANOVA

```{r, echo=FALSE}
PERMANOVA_model <- adonis2(biom_dist ~ status, data = shannon_df)
PERMANOVA_model
```


The $p$-value for PERMANOVA between the groups was `r broom::tidy(PERMANOVA_model) %>% filter(term == "status") %>% pull(p.value) %>% round(digits =2)`.

# Session Info

```{r}
sessionInfo()
```

